Is the direction of greater Granger causal influence the same as the direction of information flow?

### 2015-10-04 paper.tex
Make multiple minor edits.

In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger,dhamala2008analyzing,nolte2008robustly,korzeniewska2003determination,schippers2010mapping} etc.). In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (e.g.~\cite{Brovelli2004BetaOscillations,goebel2003investigating,deshpande2008effective}), or to acquire directed maps of functional connectivity (e.g.~\cite{friston2013analysing,goebel2003investigating,deshpande2008effective}). For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''. This raises the question: do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?

@@

Several works have outlined the procedures involved in using Granger Causality to estimate causal influences in the brain (\cite{bernasconi2000bidirectional,Brovelli2004BetaOscillations,Ding2000ShortWindow,Roebroeck2005MappingDirected,Bressler2011WienerGranger,Barnett2014MVGC,Ding2006GrangerCausality}). Here, we briefly describe how Granger Causal influence is quantified, and how it is computed in these works.

Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$. The Granger causality metric is the ratio of these errors, encapsulating the innovations that the process $\{X\}$ causally supplies to the process $\{Y\}$. Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality,quinn,jiao2013universal}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.

@@

In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$. The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations,jiao2013universal,Roebroeck2005MappingDirected}, see~\cite{Roebroeck2005MappingDirected} for an understanding of what physical constraints motivate this comparison)). Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper. We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency. It then becomes possible to also determine the brain wave frequency at which these influences occur. However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient for the purpose of our arguments.

We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison. However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations,hesse2003use,jiao2013universal}).

@@

Several objections have been raised against the use of Granger Causality in the past. We give, here, a short overview of these and describe why our objection is novel, and possibly more fundamental in nature, at least in the context of its usage in neuroscience.

\begin{enumerate}
	\item First, Granger Causality suffers from what we call the ``hidden node problem''. If two observed nodes receive causal influences from a third, latent node, then a causal influence may be detected between the observed nodes, even if they are independent of each other~\cite{pearl2009causality-hidden-node}. All nodes need to be observed, therefore, to avoid finding spurious influences.
	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction (\cite{nalatore2007mitigating,andersson2005testing}). Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influence.
	\item Third, subsampling the processes can produce misleading Granger Causal relations~\cite{gong2015discovering}. Performing Granger Causal analysis on subsampled time series can lead one to miss the causal influence. If pre-processing involves subsampling, then this should be done with care.
\end{enumerate}

It is important to note that the technical objections listed above are all deficiencies in or limitations of \emph{measurement}. They indicate that incorrect Granger Causal influences may be estimated if there is some deficiency or limitation in the measurement procedure. These objections can be resolved by taking better measurements (by sampling more nodes, with higher signal-to-noise ratio, etc.).

Our objection, on the other hand, is more fundamental. \emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, and the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence may \emph{still} not yield the correct direction of information flow. To our knowledge, the argument that Granger causal influence can be opposite to the direction of information flow is a novel one. We believe that this argument is much more serious than previous objections, at least in the context of determining the directions of information flow in neuroscientific experiments, towards understanding the computational functions of brain regions.

@@

This paper considers two experiments (introduced in Section~\ref{sec:sk}) where a transmitter $Tx$ wants to communicate a message to a receiver $Rx$ in presence of a feedback channel (in one experiment, the feedback link is noiseless, while in the other it is noisy). We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism. Provided with these measurements, the experimenter wants to estimate the direction of information flow, which in this context is the direction of flow of the message. Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that both Granger causality and Directed Information can incorrectly estimate the direction of information flow. The first experiment considers the (unrealistic) case of communication across noiseless feedback channels. The second experiment allows for noise in the feedback channel. In both cases, linear strategies inspired by the scheme of Schalkwijk and Kailath~\cite{S&K} are used.

@@

 In the first step, the transmitter sends\footnote{We assume that the power constraints are such that the scaling constant $\alpha$ in~\cite{S&K} is $1$.} $X_1 = \Theta$, which the receiver receives with added noise. The receiver sends back an estimate of $\Theta$ over the feedback link. In all subsequent iterations, the transmitter sends the receiver the error in its latest estimate.

@@

Linear strategies are known to be suboptimal for the communication problem~\cite{YoungHanKimPaper} (where $\Theta$ is a quantized random variable communicating a finite-rate message reliably), and for problems with non-classical information structures in general~\cite{Witsenhausen68}. Nevertheless, we now show that the resulting estimates $\widehat{\Theta}_i$  still converge to $\Theta$ in mean-square sense:
\begin{align*}
	i\widehat\Theta_i &= i\widehat\Theta_{i-1} + X_i + N_i \\
					  &= i\widehat\Theta_{i-1} + \Theta - \widehat\Theta_{i-1} - R_{i-1} + N_i \\
					  &= (i-1) \widehat\Theta_{i-1} + \Theta - R_{i-1} + N_i \numberthis \label{eq:theta_hat_i-noisy-once} \\
					  &\overset{(a)}{=} i\Theta + \sum_{k=1}^i N_k - \sum_{k=1}^{i-1} R_k \\
	\widehat\Theta_i  &= \Theta + \frac{1}{i} \sum_{k=1}^i N_k - \frac{1}{i} \sum_{k=1}^{i-1} R_k \numberthis \label{eq:theta_hat_i-noisy}
\end{align*}
where $(a)$ is obtained by expanding $\widehat{\Theta}_j$ recursively for $j=i-1,i-2,\ldots,2$. Therefore, the error in estimating $\Theta$ converges to $0$ in mean-square sense (\textit{i.e.}, $\widehat{\Theta}_i$ is a consistent estimate of $\Theta_i$ even in the presence of noise on the feedback link).

@@

In the forward direction, we do not explicitly compute the Granger causality ratio, but simply show that it is bounded strictly between 1 and $\infty$:
\begin{equation}
	\widehat\Theta_i = \sum_{j=1}^p{\alpha_j \widehat\Theta_{i-j}} + \epsilon_i \label{eq:ar-model-theta-theta}
\end{equation}
\begin{align*}
	\widehat\Theta_i &= \widehat\Theta_{i-1} + \frac{X_i + N_i}{i} \\
					 &= \widehat\Theta_{i-1} + \frac{\Theta - \widehat\Theta_{i-1} + N_i}{i} \\
					 &= \frac{i-1}{i} \widehat\Theta_{i-1} + \frac{\Theta}{i} + \frac{N_i}{i}
\end{align*}
which means that $0 < \frac{\sigma_N^2}{i^2} < \text{Var}(\epsilon_i) < \frac{\sigma_N^2 + \sigma_\theta^2}{i^2} < \infty$.
Further, if we try to predict $\widehat\Theta_i$ from the previous $\widehat\Theta_{i-j}$ and $X_{i-j}$:
\begin{equation}
	\widehat\Theta_i = \sum_{j=1}^{p}{\alpha_j \widehat\Theta_{i-j}} + \sum_{j=0}^{p-1}{\beta_j X_{i-j}} + \tilde\epsilon_i \label{eq:ar-model-theta-theta-x}
\end{equation}
we see that
\begin{equation*}
	\widehat\Theta_i = \widehat\Theta_{i-1} + \frac{X_i}{i} + \frac{N_i}{i}
\end{equation*}
so that $\text{Var}(\tilde\epsilon_i) = \sigma_N^2/i^2$. The Granger causality ratio, $\text{Var}(\epsilon_i) / \text{Var}(\tilde\epsilon_i)$, in the forward direction is, therefore, finite.

The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, you can always perfectly predict the transmitted symbol from the past $\widehat\Theta$'s and the history of $X$. On the other hand, you can never perfectly predict $\widehat\Theta_i$ from the past $X$'s and the history of $\widehat\Theta$.

>>>>

In order to understand how the brain performs computations, it could be useful to first understand the directions of \emph{information flow} in various parts of the brain (e.g.~\cite{blinowska2004granger, dhamala2008analyzing, nolte2008robustly, korzeniewska2003determination, schippers2010mapping} etc.). In an effort to make headway on the goals of \emph{BRAIN 2025}, several works use Granger causality (and less often, its information-theoretic generalization -- Directed Information) to understand how this information flows (e.g.~\cite{Brovelli2004BetaOscillations, goebel2003investigating, deshpande2008effective}), or to acquire directed maps of functional connectivity (e.g.~\cite{goebel2003investigating, deshpande2008effective, friston2013analysing}). For instance, in~\cite{Brovelli2004BetaOscillations}, Granger causal influences that are measured between somatosensory and motor sites are said to ``support the idea that somatosensory feedback provides information to the sensorimotor system that is used to control motor output''. This raises the question: do these directed connectivity maps, as determined by directional causal influence measures such as Granger causality, correctly identify the directions along which information flows in the brain?

@@

Several works have outlined the procedures involved in using Granger Causality to estimate causal influences in the brain (\cite{Brovelli2004BetaOscillations, bernasconi2000bidirectional, Ding2000ShortWindow, Roebroeck2005MappingDirected, Bressler2011WienerGranger, Barnett2014MVGC, Ding2006GrangerCausality}). Here, we briefly describe how Granger Causal influence is quantified, and how it is computed in these works.

Granger causality, as originally described by Granger~\cite{Granger1969InvestigatingCausal}, measures the level of causal influence that one process $\{X\}$ has on another process $\{Y\}$. The analysis compares the \emph{error in predicting} the $\{Y\}$ process based on (i) simply the past of $\{Y\}$, and (ii) based on the past of both $\{X\}$ and $\{Y\}$\footnote{A mathematical exposition of this process appears in Section~\ref{sec:gc-noiseless}}. The Granger causality metric is the ratio of these errors, encapsulating the \emph{innovations} that the process $\{X\}$ \emph{causally} supplies to the process $\{Y\}$. Many variants of Granger causality have also been developed, including a generalization -- Directed Information (see~\cite{MasseyCausality, quinn, jiao2013universal}) -- an information-theoretic quantity denoted by $I(\m{X}\to\m{Y})$. These variants form possible alternatives for estimating the direction of causal influence, but Directed Information is a generalization of many of these metrics~\cite{quinn}.

@@

In order to determine the direction of \emph{greater} causal influence, the Granger Causality metric (ratio of residual variances) from $\{X\}$ to $\{Y\}$ is often \emph{compared} to that from $\{Y\}$ to $\{X\}$. The direction of causal influence is then taken to be the direction with the greater Granger Causality metric (e.g.~\cite{Brovelli2004BetaOscillations, Roebroeck2005MappingDirected, jiao2013universal}, see~\cite{Roebroeck2005MappingDirected} for an understanding of what physical constraints motivate this comparison). Further, this direction of causal influence is interpreted to be the direction of information flow, which is the interpretation we question in this paper. We note here that Granger's original analysis does not compare this metric on forward and reverse links, and even the stronger notion of true causality (see remark~\ref{remark:gc-vs-true} in section~\ref{sec:objections-and-shortcomings}) does not involve this comparison. However, this is commonly done in practice even in areas beyond neuroscience (e.g.~\cite{Brovelli2004BetaOscillations, Roebroeck2005MappingDirected, jiao2013universal, hesse2003use}).

We also note that many of these works use a spectral version of Granger Causality, that supplies this metric as a function of frequency. It then becomes possible to also determine the brain wave frequency at which these influences occur. However, we restrict our analysis to the simpler non-spectral version of Granger Causality, since it is sufficient for the purpose of our arguments.

@@

Several objections to the use of Granger Causality have been raised in the past. We give, here, a short overview of these and describe why our objection is novel, and possibly more fundamental in nature, at least in the context of its usage in neuroscience.

\begin{enumerate}
	\item First, Granger Causality suffers from what we call the ``hidden node problem''. If two observed nodes receive causal influences from a third, latent node, then a causal influence may be detected between the observed nodes, even if they are independent of each other~\cite{pearl2009causality-hidden-node}. All nodes need to be observed, therefore, to avoid finding spurious influences.
	\item Second, if the measurements from each node are differentially affected by noise, then the predicted direction of causal influence might be opposite to the true direction (\cite{nalatore2007mitigating, andersson2005testing}). Measurements need to be relatively noiseless and precise in order to obtain the correct direction of causal influence.
	\item Third, subsampling the processes can produce misleading Granger Causal relations~\cite{gong2015discovering}. Performing Granger Causal analysis on subsampled time series can lead one to miss the causal influence. If pre-processing involves subsampling, then this should be done with care.
\end{enumerate}

It is important to note that the technical objections listed above are all deficiencies in or limitations of \emph{measurement}. They indicate that incorrect Granger Causal influences may be estimated if there is some deficiency or limitation in the measurement procedure. These objections can be resolved by taking better measurements (by sampling more nodes, using sensors with higher signal-to-noise ratio, etc.).

Our objection, on the other hand, is more fundamental. \emph{Even if} the measurements are made with infinite accuracy, and the regression coefficients associated with computing Granger Causality are precisely estimated, \emph{and} the Granger Causality metric is perfectly computed (as is the case in our counter-examples), Granger Causal influence may \emph{still} not yield the correct direction of information flow. To our knowledge, the argument that greater Granger causal influence can be opposite to the direction of information flow is a novel one. We believe that this argument is much more serious than previous objections, at least in the context of determining the directions of information flow in neuroscientific experiments, towards understanding the computational functions of brain regions.

@@

This paper considers two experiments (introduced in Section~\ref{sec:sk}) where a transmitter $Tx$ wants to communicate a message to a receiver $Rx$ in presence of a feedback channel (in one experiment, the feedback link is noiseless, while in the other it is noisy). We assume that the experimenter is able to record the transmissions of $Tx$ and $Rx$ using some probing mechanism. Provided with these measurements, the experimenter wants to estimate the direction of information flow, which in this context is the direction of flow of the message. Our results, derived in Section~\ref{sec:analytical-results} and numerically illustrated in Section~\ref{sec:numerical-results}, show that the direction of information flow can be incorrectly inferred using both Granger causality and Directed Information. The first experiment considers the (unrealistic) case of communication across noiseless feedback channels. The second experiment allows for noise in the feedback channel. In both cases, linear strategies inspired by the scheme of Schalkwijk and Kailath~\cite{S&K} are used.

@@

 In the first step, the transmitter sends\footnote{We assume that the power constraints are such that the scaling constant `$\alpha$' in~\cite{S&K} is $1$.} $X_1 = \Theta$, which the receiver receives with added noise. The receiver sends back an estimate of $\Theta$ over the feedback link. In all subsequent iterations, the transmitter sends the receiver the error in its latest estimate.

@@

Linear strategies are known to be suboptimal for this communication problem~\cite{YoungHanKimPaper} (where $\Theta$ is a quantized random variable communicating a finite-rate message reliably), and for problems with non-classical information structures in general~\cite{Witsenhausen68}. Nevertheless, we now show that the resulting estimates $\widehat{\Theta}_i$  still converge to $\Theta$ in mean-square sense:
\begin{align*}
	i\widehat\Theta_i &= i\widehat\Theta_{i-1} + X_i + N_i \\
					  &= i\widehat\Theta_{i-1} + \Theta - \widehat\Theta_{i-1} - R_{i-1} + N_i \\
					  &= (i-1) \widehat\Theta_{i-1} + \Theta - R_{i-1} + N_i \numberthis \label{eq:theta_hat_i-noisy-once} \\
					  &\overset{(a)}{=} i\Theta + \sum_{k=1}^i N_k - \sum_{k=1}^{i-1} R_k \\
	\widehat\Theta_i  &= \Theta + \frac{1}{i} \sum_{k=1}^i N_k - \frac{1}{i} \sum_{k=1}^{i-1} R_k \numberthis \label{eq:theta_hat_i-noisy}
\end{align*}
where $(a)$ is obtained by expanding $\widehat{\Theta}_j$ recursively for $j=i-1,i-2,\ldots,2$. Therefore, the error in estimating $\Theta$ converges to $0$ in mean-square sense (\textit{i.e.}, $\widehat{\Theta}_i$ is a consistent estimate of $\Theta_i$ even in the presence of noise on the feedback link).

@@

In the forward direction, we do not explicitly compute the Granger causality ratio, but simply show that it is bounded strictly between 1 and $\infty$:
\begin{equation}
	\widehat\Theta_i = \sum_{j=1}^p{\alpha_j \widehat\Theta_{i-j}} + \epsilon_i \label{eq:ar-model-theta-theta}
\end{equation}
\begin{align*}
	\widehat\Theta_i &= \widehat\Theta_{i-1} + \frac{X_i + N_i}{i} \\
					 &= \widehat\Theta_{i-1} + \frac{\Theta - \widehat\Theta_{i-1} + N_i}{i} \\
					 &= \frac{i-1}{i} \widehat\Theta_{i-1} + \frac{\Theta}{i} + \frac{N_i}{i}
\end{align*}
which means that $0 < \frac{\sigma_N^2}{i^2} < \text{Var}(\epsilon_i) < \frac{\sigma_N^2 + \sigma_\theta^2}{i^2} < \infty$, since the past of $\widehat{\Theta}$ cannot be used to explain $N_i$.
Further, if we try to predict $\widehat\Theta_i$ from the previous $\widehat\Theta_{i-j}$ and $X_{i-j}$:
\begin{equation}
	\widehat\Theta_i = \sum_{j=1}^{p}{\alpha_j \widehat\Theta_{i-j}} + \sum_{j=0}^{p-1}{\beta_j X_{i-j}} + \tilde\epsilon_i \label{eq:ar-model-theta-theta-x}
\end{equation}
we see that
\begin{equation*}
	\widehat\Theta_i = \widehat\Theta_{i-1} + \frac{X_i}{i} + \frac{N_i}{i}
\end{equation*}
so that $\text{Var}(\tilde\epsilon_i) = \sigma_N^2/i^2$. The Granger causality ratio, $\text{Var}(\epsilon_i) / \text{Var}(\tilde\epsilon_i)$, in the forward direction is, therefore, finite.

The intuitive argument for why this is happening might go as follows: since the feedback link is noiseless, one can always perfectly predict the transmitted symbol from the past $\widehat\Theta$'s and the history of $X$. On the other hand, one can never perfectly predict $\widehat\Theta_i$ from the past $X$'s and the history of $\widehat\Theta$.
